Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2013-01-15 paper.tex
some simpifications

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is not consistent, but converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

We assume that time offset estimation has been performed and that $L$ noisy $M$-PSK symbols are observed by the receiver.  The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

@@

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Swelden considered only the noncoherent setting, but we show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\sabs{\hat{a} - a_0}$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\sabs{\langle\hat{\theta} - \theta_0\rangle_\pi}$ and $\sabs{\hat{\rho} - \rho_0}$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\hat{\theta} - \theta_0\rangle_\pi$ rather than $\hat{\theta} - \theta_0$ is the appropriate measure of error for the phase parameter.

It is possible to generalise the results we present here to allow data symbols with varying constellation size, i.e. varying $M$.  For example, one might give more importance to certain data symbols and use BPSK ($M=2$) for these, but QPSK ($M=4$) for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  To keep our ideas and notation focused we don't consider this further here.

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some reasonably general assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\hat{\theta} - \theta_0\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\hat{\theta} - \theta_0\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:SS}
\end{align}
where $*$ denotes complex conjugate.  The minimiser of $SS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain $SS$ minimised with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the $d_i$.  The least squares estimators of the data symbols are the minimisers of~\eqref{eq:SSdatasymbols}.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$ where $\rho$ is a nonnegative real.  Now,
\begin{align}
SS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:SSallparams}
\end{align}
We have slightly abused notation here by reusing $SS$. This should not cause confusion as $SS(a, \{d_i, i \in D\})$, $SS(\rho, \theta, \{d_i, i \in D\})$, and $SS(\{d_i, i \in D\})$ are easily told apart by their arguments.  For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

We only require to consider $\theta$ in the interval $[0, 2\pi)$.  Consider how $\hat{d}_i(\theta)$ changes as $\theta$ varies from $0$ to $2\pi$.  Let $b_i = \hat{d}_i(0)$ and let 
\[
z_i = \angle{y_i} - \hat{u}_i(0) = \angle{y_i} - \round{\angle{y_i}}
\]
be the phase difference between the received symbol $y_i$ and the hard decision resulting when $\theta = 0$, i.e. $\round{\angle{y_i}}$.  Then,
\begin{equation}\label{eq:uicombos}
\hat{d}_i(\theta) = 
\begin{cases}
b_i, &  0 \leq \theta < z_i + \frac{\pi}{M} \\
b_i e^{-j2\pi/M}, & z_i + \frac{\pi}{M} \leq \theta < z_i + \frac{3\pi}{M} \\ 
\vdots & \\
b_i e^{-j2\pi k /M}, & z_i + \frac{\pi(2k - 1)}{M} \leq \theta < z_i + \frac{\pi(2k + 1)}{M}  \\ 
\vdots & \\
b_i e^{-j2\pi} = b_i, &  z_i + \frac{\pi(2M - 1)}{M} \leq \theta < 2\pi. \\
\end{cases}
\end{equation}

Let 
\[
f(\theta) = \{ \hat{d}_i(\theta), i \in D \}
\]
be a function mapping the interval $[0, 2\pi)$ to a sequence of $M$-PSK symbols indexed by the elements of $D$.  Observe that $f(\theta)$ is piecewise continuous.  The subintervals of $[0, 2\pi)$ over which $f(\theta)$ remains constant are determined by the values of $\{z_i, i \in D\}$.  Let
\[
S = \{ f(\theta) \mid \theta \in [0, 2 \pi) \}
\]
be the set of all sequences $f(\theta)$ as $\theta$ varies from $0$ to $2\pi$.  If $\hat{\theta}$ is the least squares estimator of the phase then $S$ contains the sequence $\{ \hat{d}_i(\hat{\theta}), i \in D \}$ corresponding to the least squares estimator of the data symbols, i.e., $S$ contains the minimiser of~\eqref{eq:SSdatasymbols}.  Observe from~\eqref{eq:uicombos} that there are at most $M\abs{D}$ sequences in $S$, because there are $M$ distinct values of $d_i(\theta)$ for each $i \in D$ as $\theta$ varies from $0$ to $2\pi$.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = e^{j2\pi/M} - 1$.  In general,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a $O(1)$ arithmetic operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in $O(1)$ operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of $a_0$ is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Pseudocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindices}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is not consistent and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  %Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  
The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

@@

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered only the noncoherent setting, but we show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the data symbols is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\sabs{\hat{a} - a_0}$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\sabs{\langle\hat{\theta} - \theta_0\rangle_\pi}$ and $\sabs{\hat{\rho} - \rho_0}$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\hat{\theta} - \theta_0\rangle_\pi$ rather than $\hat{\theta} - \theta_0$ is the appropriate measure of error for the phase parameter.

%It is possible to generalise the results we present here to allow data symbols with varying constellation size, i.e. varying $M$.  For example, one might give more importance to certain data symbols and use BPSK ($M=2$) for these, but QPSK ($M=4$) for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  To keep our ideas and notation focused we don't consider this further here.

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under reasonably general assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\hat{\theta} - \theta_0\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\hat{\theta} - \theta_0\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:SS}
\end{align}
where $*$ denotes complex conjugate.  The minimiser of $SS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain $SS$ minimised with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the $d_i$.  The least squares estimators of the data symbols are the minimisers of~\eqref{eq:SSdatasymbols}.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$ where $\rho > 0$ is real.  Now,
\begin{align}
SS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:SSallparams}
\end{align}
We have slightly abused notation here by reusing $SS$. This should not cause confusion as $SS(a, \{d_i, i \in D\})$, $SS(\rho, \theta, \{d_i, i \in D\})$, and $SS(\{d_i, i \in D\})$ are easily told apart by their arguments.  For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

We only require to consider $\theta$ in the interval $[0, 2\pi)$.  Consider how $\hat{d}_i(\theta)$ changes as $\theta$ varies from $0$ to $2\pi$.  Let $b_i = \hat{d}_i(0)$ and let 
\[
z_i = \angle{y_i} - \hat{u}_i(0) = \angle{y_i} - \round{\angle{y_i}}
\]
%be the phase difference between the received symbol $y_i$ and the hard decision resulting when $\theta = 0$, i.e. $\round{\angle{y_i}}$.  
Then,
\begin{equation}\label{eq:uicombos}
\hat{d}_i(\theta) = 
\begin{cases}
b_i, &  0 \leq \theta < z_i + \frac{\pi}{M} \\
b_i e^{-j2\pi/M}, & z_i + \frac{\pi}{M} \leq \theta < z_i + \frac{3\pi}{M} \\ 
\vdots & \\
b_i e^{-j2\pi k /M}, & z_i + \frac{\pi(2k - 1)}{M} \leq \theta < z_i + \frac{\pi(2k + 1)}{M}  \\ 
\vdots & \\
b_i e^{-j2\pi} = b_i, &  z_i + \frac{\pi(2M - 1)}{M} \leq \theta < 2\pi. \\
\end{cases}
\end{equation}

Let 
\[
f(\theta) = \{ \hat{d}_i(\theta), i \in D \}
\]
be a function mapping the interval $[0, 2\pi)$ to a sequence of $M$-PSK symbols indexed by the elements of $D$.  Observe that $f(\theta)$ is piecewise continuous.  The subintervals of $[0, 2\pi)$ over which $f(\theta)$ remains constant are determined by the values of $\{z_i, i \in D\}$.  Let
\[
S = \{ f(\theta) \mid \theta \in [0, 2 \pi) \}
\]
be the set of all sequences $f(\theta)$ as $\theta$ varies from $0$ to $2\pi$.  If $\hat{\theta}$ is the least squares estimator of the phase then $S$ contains the sequence $\{ \hat{d}_i(\hat{\theta}), i \in D \}$ corresponding to the least squares estimator of the data symbols, i.e., $S$ contains the minimiser of~\eqref{eq:SSdatasymbols}.  Observe from~\eqref{eq:uicombos} that there are at most $M\abs{D}$ sequences in $S$, because there are $M$ distinct values of $\hat{d}_i(\theta)$ for each $i \in D$ as $\theta$ varies from $0$ to $2\pi$.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = e^{j2\pi/M} - 1$.  In general,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in $O(1)$ arithmetic operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in $O(1)$ operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of $a_0$ is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Pseudocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindices}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.
